Formal Proof of the Relative Consistency of ZF$+\neg$AC with Isabelle/ZF

### 2024-12-24 paper/main.tex
updated

\documentclass{report}
\usepackage{graphicx} 
\usepackage{luacode}  
\usepackage{ascmac}
\usepackage{comment}
\usepackage{amsthm}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{xcolor}
\usepackage{authblk}

@@

% 参考文献
\usepackage[style=numeric,backend=biber]{biblatex}
\addbibresource{main.bib}

@@

\subsubsection{Proof Assistants}

Proof assistants are software tools designed to formally verify mathematical proofs, programs, and more. 
These systems are widely used in various fields, including mathematics, cryptography, software verification, and hardware design.
Operating on the basis of formal logic, they enable users to interactively construct proofs, which are then checked by the system.
Well-known examples of proof assistants include Isabelle, Coq, Lean, and HOL Light,
each offering unique features and strengths.

One of the key advantages of proof assistants is their ability to guarantee the correctness of proofs. 
In traditional, informal mathematical proofs, such as those created with pen and paper, 
errors can easily occur due to human oversight or carelessness. 
In contrast, by using proof assistants, every logical step is rigorously verified by a machine, 
ensuring that the proof is formally correct.

For example, the Four Color Theorem \cite{four_color} and Kepler's Conjecture \cite{kepler} are well-known cases of theorems formally proven using proof assistants. 
These large-scale proofs were difficult for humans to verify manually, but their correctness was assured through formal verification.
Similarly, in systems where safety is critical, formal verification of programs can significantly enhance their reliability. 
Notable examples of formally verified systems include the CompCert C compiler \cite{compcert} and the seL4 microkernel \cite{seL4}.

@@

\subsubsection{Axiomatic Set Theory}

@@

\subsubsection{Relative Consistency and Independence}

@@

\subsubsection{Forcing}

Forcing is a technique first introduced by Cohen in 1963 \cite{cohen} to prove the independence of AC in ZF 
and the independence of CH in ZFC. 
It is a powerful method for exploring models of set theory, 
and it has been further refined by other researchers over time, 
leading to many significant results being established through its use. 
Forcing continues to play a crucial role in contemporary research in axiomatic set theory.

@@

\subsubsection{Previous Research on the Formalization of Axiomatic Set Theory}

@@

Paulson et al. formalized an extensive part of ZF set theory 
\cite{paulson_datatype_impl,paulson_reflection,paulson_AC_consistency,paulson_cardinal_AC,paulson_datatype}, 
including cardinal arithmetic, relativization, the reflection theorem, features for handling inductive definitions, and the relative consistency of AC with ZF.
The proof of the relative consistency was achieved by constructing Gödel's constructible universe.

Building on these results, Gunther et al. formalized forcing and a proof of the independence of CH \cite{gunther_forcing,gunther_independence} in Isabelle/ZF.
In these formalizations, the countable transitive model (c.t.m.) approach was used, following Kunen's book \cite{kunen2011}.

@@

Although these results have been formalized, 
the relative consistency of $\neg$AC with ZF has not yet been formalized. 
It can be proven by forcing, but the proof involves complexities that cannot be
achieved by simply modifying the proof for CH.

\subsubsection{Our Contribution and Approach}

In this paper, we formalize the proof of the relative consistency of $\neg$AC with ZF, which has not yet been formalized, using Isabelle/ZF.

The primary reason we chose Isabelle/ZF for this formalization is that Isabelle/ZF is a mature proof assistant for ZF set theory,
in particular, the formalization by Gunther et al. \cite{gunther_forcing} is a major advantage for this study.
Although there is also a formalization of forcing in Lean 3 by Han and van Doorn \cite{flypitch}, development for Lean 3 has already ended.

To use Gunther et al.'s formalization, we adopted the c.t.m.\  approach for our proof, as in their independence proof of CH.
Specifically, we assumed the existence of a c.t.m.\  of ZF and constructed a model of ZF+$\neg$AC by forcing.
This model is known as the basic Cohen model, which is a symmetric extension of assumed c.t.m.\ 
Our proof is based on Karagila's lecture note \cite{karagila} and Jech's book \cite{jech_set_theory,jech_AC},
as these resources align well with this approach.

The c.t.m.\  approach is a well-established method for relative consistency proofs in axiomatic set theory.
The relative consistency of $\neg$AC with ZF means that if ZF is consistent, then ZF+$\neg$AC is also consistent.
Assuming the consistency of ZF implies that a model of ZF exists, but strictly speaking, the existence of a c.t.m.\  cannot be derived from this assumption.
However, as explained in \cref{subsection:ctm}, we can prove the relative consistency of $\neg$AC with ZF if 
we can construct a model of ZF+$\neg$AC from a c.t.m.\  of ZF.
This kind of reasoning always arises in the c.t.m.\  approach, and ideally, 
we would like to formalize it as well, but this has not been achieved.

This work contributes to the formalization of axiomatic set theory and serves
as a new example of the formalization using forcing, which is a crucial tool in set theory.
It also may provide insights into how the formalization of axiomatic set theory could be advanced.

\subsubsection{Synopsis}
The remainder of this thesis is outlined as follows.
In Chapter 2, we first explain set-theoretic concepts in Section 2.1.
Section 2.2 outlines an informal proof of the relative consistency of $\neg$AC with ZF, which is formalized in Chapter 3.
In Section 2.3, we provide an overview of Isabelle/ZF and highlight key results from prior work that are relevant to this study.
Chapter 3 presents the formalization of the relative consistency proof of ZF+$\neg$AC, building on the contents of Chapter 2.
Finally, Chapter 4 concludes the thesis.

\subsubsection{Repository}
Our source code is available at: \url{https://github.com/tarakojo/ZF_notAC}

@@

\section{Set-Theoretic Preliminaries}
In this section, we introduce the concepts of set theory used in the formalization of this study.
We discuss in ZFC, within which we explore ZF itself. 
Basically, our definition and theorems combine Kunen \cite{kunen2011}, karagila's lecture note \cite{karagila}, and Jech's book \cite{jech_AC}
adapted to the formalized form in Isabelle/ZF.

@@

\subsection{The c.t.m.\  Approach}\label{subsection:ctm}
In this section, we explain the validity of the c.t.m.\  approach, 
which is a method for proving the relative consistency and independence in axiomatic set theory.

The relative consistency of $\neg$AC with ZF means that if ZF is consistent, then ZF+$\neg$AC is also consistent.
When ZF is consistent, we can prove, by the completeness theorem, that ZF has a model.
However, the existence of a c.t.m.\  of ZF cannot be derived from the consistency of ZF.
The statement that ZF has a c.t.m.\  is stronger than the consistency of ZF.
Nonetheless, by assuming the existence of a c.t.m.\  of ZF and then constructing a model of ZF+$\neg$AC by forcing, 
we can prove the relative consistency of $\neg$AC with ZF.
To explain this, we introduce some concepts,
following Jech \cite{jech_set_theory}, chapter 6 and 12.

@@

Here, we use this to explain the validity of the c.t.m.\ approach.
Note that this discussion is not limited to the relative consistency proof of $\neg$AC with ZF;
a similar argument can be applied to other relative consistency proofs using the c.t.m.\ approach.

First, let us assume the existence of a c.t.m.\  of ZF and 
suppose that we can prove the existence of a c.t.m.\ of ZF+$\neg$AC by forcing.
It should also be possible to prove the following by modifying the proof.

@@

\subsection{Isabelle/ZF}
Isabelle \cite{paulson_1986} is a proof assistant developed by Paulson, used for formalizing mathematical proofs and verifying programs. 
Since its release in 1986, it has continued to be developed by many researchers.

@@

\subsection{Important Results from Prior Work}

\subsubsection{Recursive Data Types and Functions}
Paulson \cite{paulson_datatype_impl} introduced \texttt{"datatype"} to recursive data types within ZF set theory, 
and "primrec" to define functions on top of these data types. 
This made it possible to define data structures such as lists within ZF set theory.
For example, the set of all lists of elements from a set $A$ is defined in a library of Isabelle/ZF as follows:
\begin{isaframe}
\begin{minted}[breaklines]{isabelle}
consts
    list       :: "i⇒i"

@@

\chapter{Formalization of the Proof} 
In this chapter, we present the formalization of the relative consistency proof of $\neg$AC with ZF in Isabelle/ZF.
We follow the outline presented in \cref{sec:outline}.
We choose \emph{ZF-Constructible} by Paulson \cite{paulson_AC_consistency} as Isabelle theory 
\footnote{Note that this does not mean assuming $V=L$.}
, which includes the ZF axioms along with useful definitions and lemmas built on top of them.
We also reuse the formalization of forcing by Gunther et al. \cite{gunther_forcing}.

@@

\subsubsection{$\mathbb{P}$-names}
We begin our work within the locale \texttt{"forcing\_data"} by Gunther et al. \cite{gunther_forcing}, 
in which a c.t.m.\  \texttt{"M"} and a notion of forcing \texttt{"P"} are fixed.
The relation $\leq_{\mathbb{P}}$ is denoted by \texttt{"leq"} and 
a maximum element $1_{\mathbb{P}}$ is denoted by \texttt{"one"}.

@@

Where \texttt{"val(G, x)"} is the interpretation $x^G$ formalized by Gunther et al. \cite{gunther_forcing}.

We declare a locale \texttt{"M\_symmetric\_system\_G\_generic"},
which combines the locale \texttt{"M\_symmetric\_system"} and the locale \texttt{"G\_generic"}.
\texttt{"G\_generic"} is a locale, defined by Gunther et al. \cite{gunther_forcing}, fixes a generic filter \texttt{"G"} on \texttt{"P"}.

@@

We utilized the automated formula synethisis method by Gunther et al. \cite{gunther_forcing};
however, it could not applied to complex formulas, 
and re-verifying the properties of the derived formulas in a more usable form remained a repititive task.

@@

In this section, we prove that symmetric extensions are models of ZF.
Initially, we attempted to prove this based on its proof in Karagila's lecture note \cite{karagila},
but this approach was unsuccessful.
Instead, we prove this by using the relativized forcing relation $\Vdash_{\mathrm{HS}}$.
Karagila's approach is applicable to a more general situation,
without using forcing or generic and symmetric extensions,
whereas our approach is more specific to the context of symmetric extensions.

\subsubsection{Unsuccessful Approach}

In Karagila's lecture note \cite{karagila},
the fact that symmetric extensions are models of ZF is proven by showing that 
they are internal models of the conrresponding generic extension
generated by the same notion of forcing and generic filter,
using the following proposition:

@@

We could prove that the symmetric extension \texttt{"SymExt(G)"} is transitive, almost universal and satisfies $\Delta_0$-separation.
However, we could not prove that \texttt{"SymExt(G)"} is actually a class in \texttt{"GenExt(G)"}.
To do this, we needed to explicitly construct a formula that represents the set of all names \texttt{"P\_names"} on the ground model \texttt{"M"} in \texttt{"GenExt(G)"}.
However, we were unable to construct such a formula.

\subsubsection{The Relativized Forcing Relation $\Vdash_{\mathrm{HS}}$}
Instead, we formalize the relativized forcing relation $\Vdash_{\mathrm{HS}}$ and prove that symmetric extensions are models of ZF using it.
By modifying the formalization of forcing by Gunther et al. \cite{gunther_forcing}, 
we were able to define this relation and prove its properties relatively easily.

@@

Where \texttt{"wellordered(C, A, r)"} denotes that 
the relation \texttt{"r"} is a well-ordering of the set \texttt{"A"} in the class \texttt{"C"}, 
and \texttt{"\#\#"} is an operator that converts a set to a class.
These are defined by Paulson \cite{paulson_AC_consistency}.
Specifically, the definition of \texttt{"wellordered"} is as follows
\footnote{
Note that we can show that 
if \texttt{"wellordered(N, A, r)"} holds for an element \texttt{"A"} in a class \texttt{"N"}, which is a model of ZF,
then \texttt{"r"} is irreflexive and assymetric, i.e., a strict total order.
}:
\begin{isaframe}
\begin{minted}[breaklines]{isabelle}
definition transitive_rel :: "[i⇒o,i,i]⇒o" where
    "transitive_rel(M,A,r) ≡ 
        ∀x[M]. x∈A ⟶ (∀y[M]. y∈A ⟶ (∀z[M]. z∈A ⟶ 
                          ⟨x,y⟩∈r ⟶ ⟨y,z⟩∈r ⟶ ⟨x,z⟩∈r))"
definition linear_rel :: "[i⇒o,i,i]⇒o" where
    "linear_rel(M,A,r) ≡ 
        ∀x[M]. x∈A ⟶ (∀y[M]. y∈A ⟶ ⟨x,y⟩∈r | x=y | ⟨y,x⟩∈r)"

@@

\chapter{Conclusion}

We formalized the relative consistency proof of $\neg$AC with ZF
on the basis of the work by Paulson \cite{paulson_AC_consistency} and Gunther et al. \cite{gunther_forcing}.

@@

In our proof, we needed to construct internalized formulas 
representing most of the concepts that belong to the ground model. 
This was to verify that the defined sets truly belong to the ground model 
and to enable tha application of the separation and replacement axioms of the ground model for formulas that include these concepts.

For constructing these formulasm we utilized the sythesis method by Gunther et al. \cite{gunther_forcing};
however, this method could not be applied to complex formulas.
When applied, we also needed to reprove properties of derived formulas in a more usable form.

@@

\subsubsection{Future Work}

@@

Additionally, we used the c.t.m.\ approach, and the formalized argument only extends to the construction of a model of ZF+$\neg$AC,
assuming the existence of a c.t.m.\ of ZF.
We did not formalize the justification of the c.t.m.\ approach itself.
Formalizing this part would be a interesting future work.
Ultimately, it would be desirable to derive the relative consistency proof automatically from the assumption of the existence of a c.t.m.; 
however, achieving this within Isabelle/ZF, which is based on first-order logic, seems infeasible.

In addition, formalizing the connections between our approach and alternative approaches to forcing, 
such as Boolean-valued models or syntactic methods, is another topic to explore in the future.

\printbibliography
\end{document}

>>>>

\documentclass{report}
\usepackage{graphicx} 
\usepackage{luacode}  
\usepackage{ascmac}
\usepackage{comment}
\usepackage{amsthm}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{xcolor}
\usepackage{authblk}
\usepackage{url}

@@

% 参考文献
\usepackage[backend=biber]{biblatex}
\addbibresource{main.bib}

@@

\paragraph{Proof Assistants}

Proof assistants are software tools designed to formally verify mathematical proofs, programs, and more. 
These systems are widely used in various fields, including mathematics \cite{four_color,kepler}, cryptography \cite{EasyCrypt_AWS,EasyCrypt_zero_knowledge}, 
hardware verification \cite{Coq_hardware_verification,Coq_hardware_verification_2},
and program verification \cite{survey_proof_assistants_program_verification}. 
Operating on the basis of formal logic, they enable users to interactively construct proofs, which are then checked by the system.
Well-known examples of proof assistants include Isabelle \cite{isabelle_toppage}, Coq \cite{coq_toppage}, Lean \cite{lean_toppage} Agda \cite{agda_toppage}, and HOL Light \cite{hol_light_toppage},
each offering unique features and strengths \cite{survey_theorem_provers}.

One of the key advantages of proof assistants is their ability to guarantee the correctness of proofs. 
In manually constructed proofs, errors can easily occur due to human mistakes, 
and even when the proof is correct, it can be difficult to verify its correctness.
Proof assistants provide a solution to these problems by rigorously checking each logical step of given proof,
ensuring its correctness.

In particular, these systems are effective in verifying complex proofs.
For example, the Four Color Theorem \cite{four_color} and Kepler's Conjecture \cite{kepler} are well-known cases of theorems formally proven using proof assistants. 
These large-scale proofs were difficult for humans to verify manually, but their correctness was assured through formal verification.
Similarly, in systems where safety is critical, formal verification of programs can significantly enhance their reliability. 
Notable examples of formally verified systems include the CompCert C compiler \cite{compcert} and the seL4 microkernel \cite{seL4}.

@@

\paragraph{Axiomatic Set Theory}

@@

While this paper provides a brief explanation of axiomatic set theory, 
a detailed exposition can be found in standard textbooks such as \cite{kunen2011,jech_set_theory}.

\paragraph{Relative Consistency and Independence}

@@

\paragraph{Forcing}

Forcing is a technique first introduced by Cohen \cite{cohen} to prove the independence of AC in ZF 
and the independence of CH in ZFC. 
It is a powerful method for exploring models of set theory, 
and it has been further refined by other researchers over time, 
leading to many significant results being established through its use. 
Forcing continues to play a crucial role in contemporary research in axiomatic set theory.

@@

\paragraph{Previous Research on the Formalization of Axiomatic Set Theory}

@@

Paulson et al. formalized an extensive part of ZF set theory,
including cardinal arithmetic \cite{paulson_cardinal_AC}, relativization \cite{paulson_AC_consistency}, the reflection theorem \cite{paulson_reflection}, 
features for handling inductive definitions \cite{paulson_datatype_impl}, and the relative consistency of AC with ZF \cite{paulson_AC_consistency}
The proof of the relative consistency was achieved by constructing Gödel's constructible universe.

Building on these results, Gunther et al. formalized forcing and a proof of the independence of CH \cite{gunther_forcing,gunther_independence} in Isabelle/ZF.
In these formalizations, they used the countable transitive model (c.t.m.) approach, 
which we also adopt in this paper.

@@

Although these results have been formalized, 
the relative consistency of $\neg$AC with ZF has not yet been formalized.

\paragraph{Our Contribution and Approach}

In this paper, we formalize the proof of the relative consistency of $\neg$AC with ZF using Isabelle/ZF.

This work contributes to the formalization of axiomatic set theory 
and serves as a new example of the formalization using forcing, which is a crucial tool in set theory.
It also highlights parts of the process that require significant effort 
and provides insights into how future formalizations could be made more efficient.

The primary reason we chose Isabelle/ZF for this formalization is that Isabelle/ZF is a mature proof assistant for ZF set theory,
in particular, the formalization by \cite{gunther_forcing} is a major advantage for this study.
Although there is also a formalization of forcing in Lean 3 by Han and van Doorn \cite{flypitch}, development for Lean 3 has already ended.

We adopted the c.t.m.\  approach for our proof to use Gunther et al.'s formalization, as in their independence proof of CH.
Specifically, we show th relative consistency of $\neg$AC with ZF 
by assuming the existence of a c.t.m.\  of ZF and constructing a model of ZF+$\neg$AC by forcing.
This model is known as the basic Cohen model, which is a symmetric extension of assumed c.t.m.\ 
Constructing the basic Cohen model by forcing is 
one of traditional methods for proving the relative consistency of $\neg$AC with ZF.

Our proof is based on Karagila's lecture note \cite{karagila} and Jech's book \cite{jech_AC},
as these resources align well with the c.t.m.\  approach.
However, some parts of the proof required modification to fit the formalization in Isabelle/ZF.

The c.t.m.\ approach is a well-established method for relative consistency proofs in axiomatic set theory.
This approach will be explained in detail in \cref{subsection:ctm}.
The discussion regarding the validity of the c.t.m.\ approach, 
which is often omitted or only briefly mentioned in paper-based proofs, has not yet been formalized, and its formalization is desirable.
Further details about this will be provided in \cref{chap:conclusion}.

\paragraph{Synopsis}
The remainder of this thesis is outlined as follows.
In Chapter 2, we first explain set-theoretic concepts in Section 2.1.
Section 2.2 outlines an informal proof (i.e., a proof on paper) of the relative consistency of $\neg$AC with ZF, which is formalized in Chapter 3.
In Section 2.3, we provide an overview of Isabelle/ZF and highlight key results from prior work that are relevant to this study.
Chapter 3 presents the formalization of the relative consistency proof of ZF+$\neg$AC, building on the contents of Chapter 2.
Finally, Chapter 4 concludes the thesis.

\paragraph{Repository}
Our source code is available at: \url{https://github.com/tarakojo/ZF_notAC}

@@

\section{Set-Theoretic Preliminaries}
In this section, we introduce the concepts of set theory used in the formalization of this study.
We discuss in ZFC, within which we explore ZF itself. 
Basically, our definition and theorems combine \cite{kunen2011}, \cite{karagila}, and \cite{jech_AC}
adapted to the formalized form in Isabelle/ZF.

@@

\subsection{The c.t.m.\  Approach}\label{subsection:ctm}
This section explains the c.t.m.\ approach.
This approach is a well-established method for proving relative consistency in axiomatic set theory,
as it is covered in \cite{kunen2011}.

Specifically, the approach assumes the existence of a countable transitive model (c.t.m.) 
of an axiom system $S$ of set theory and show the existence of a c.t.m.\ for another axiom system $T$, 
thereby proving the relative consistency of $T$ with respect to $S$.
This approach is useful because it allows discussions to proceed without worrying about the complicated details that can arise.

It should be noted that discussions are needed to confirm the validity of this approach.
By definition, the relative consistency of $T$ with respect to $S$ means that if $S$ is consistent, then $T$ is also consistent.
Assuming that $S$ is consistent, by the completeness theorem, we can prove that $S$ has a model.
However, the existence of a c.t.m.\  of $S$ cannot be necessarily derived from the consistency of $S$.

Nonetheless, the c.t.m.\  approach is valid for proving the relative consistency.
The following presents a discussion on the validity of the approach.
In accordance with the content of this paper, we proceed with the discussion by letting $S$ be ZF and $T$ be ZF + $\neg$AC.
However, the discussion below is also applicable to other cases.

First, we introduce some concepts,
following Jech \cite{jech_set_theory}, chapter 6 and 12.

@@

Here, we use this to explain the validity of the c.t.m.\ approach.
First, let us assume the existence of a c.t.m.\  of ZF and 
suppose that we can prove the existence of a c.t.m.\ of ZF+$\neg$AC by forcing.
It should also be possible to prove the following by modifying the proof.

@@

\subsection{Isabelle and Isabelle/ZF}
Isabelle is a proof assistant developed \cite{paulson_1986}, used for formalizing mathematical proofs and verifying programs. 
Since its release in 1986, it has continued to be developed by many researchers.

@@

\subsection{Important Results from Prior Work in Isabelle/ZF}

\subsubsection{Recursive Data Types and Functions}
\cite{paulson_datatype_impl} introduced \texttt{"datatype"} to recursive data types within ZF set theory, 
and "primrec" to define functions on top of these data types. 
This made it possible to define data structures such as lists within ZF set theory.
For example, the set of all lists of elements from a set $A$ is defined in a library of Isabelle/ZF as follows:
\begin{isaframe}
\begin{minted}[breaklines]{isabelle}
consts
    list       :: "i⇒i"

@@

\chapter{Formalization of the Proof} 
In this chapter, we present the formalization of the relative consistency proof of $\neg$AC with ZF in Isabelle/ZF.
We follow the outline presented in \cref{sec:outline}.
We choose \emph{ZF-Constructible} by \cite{paulson_AC_consistency} as Isabelle theory 
\footnote{Note that this does not mean assuming $V=L$.}
, which includes the ZF axioms along with useful definitions and lemmas built on top of them.
We also reuse the formalization of forcing by \cite{gunther_forcing}.

@@

\subsubsection{$\mathbb{P}$-names}
We begin our work within the locale \texttt{"forcing\_data"} by \cite{gunther_forcing}, 
in which a c.t.m.\  \texttt{"M"} and a notion of forcing \texttt{"P"} are fixed.
The relation $\leq_{\mathbb{P}}$ is denoted by \texttt{"leq"} and 
a maximum element $1_{\mathbb{P}}$ is denoted by \texttt{"one"}.

@@

Where \texttt{"val(G, x)"} is the interpretation $x^G$ formalized by \cite{gunther_forcing}.

We declare a locale \texttt{"M\_symmetric\_system\_G\_generic"},
which combines the locale \texttt{"M\_symmetric\_system"} and the locale \texttt{"G\_generic"}.
\texttt{"G\_generic"} is a locale, defined by \cite{gunther_forcing}, fixes a generic filter \texttt{"G"} on \texttt{"P"}.

@@

We utilized the automated formula synethisis method by \cite{gunther_forcing};
however, it could not applied to complex formulas, 
and re-verifying the properties of the derived formulas in a more usable form remained a repititive task.

@@

In this section, we prove that symmetric extensions are models of ZF.
Initially, we attempted to formalize this based on its proof in \cite[Theorem 10.10]{karagila},
but we found that additional work was required to formalize parts that are omitted in the note.
Therefore, we decided to prove this by using the relativized forcing relation $\Vdash_{\mathrm{HS}}$.
Initial approach is a traditional method for proving that symmetric extensions are models of ZF,
and applicable to a more general situation since 
it does not rely on forcing or generic and symmetric extensions.
whereas our approach is more specific to the context of symmetric extensions.

\subsubsection{Traditional Approach}

In \cite[Theorem 10.10]{karagila} and \cite{jech_set_theory},
the fact that symmetric extensions are models of ZF is proven by showing that 
they are internal models of the conrresponding generic extension
generated by the same notion of forcing and generic filter,
using the following proposition:

@@

We could prove that the symmetric extension \texttt{"SymExt(G)"} is transitive, almost universal and satisfies $\Delta_0$-separation.
However, 
we found that proving that \texttt{"SymExt(G)"} is actually a class in \texttt{"GenExt(G)"} is not straightforward in Isabelle/ZF.
To do this, we needed to explicitly construct a formula that represents the set of all names \texttt{"P\_names"} on the ground model \texttt{"M"} in \texttt{"GenExt(G)"}.
However, this required additional work, which would likely involve complex discussions.
Therefore, we decided to adopt a different approach.

\subsubsection{Our Approach : The Relativized Forcing Relation $\Vdash_{\mathrm{HS}}$}
Instead, we formalize the relativized forcing relation $\Vdash_{\mathrm{HS}}$ and prove that symmetric extensions are models of ZF using it.
By modifying the formalization of forcing by \cite{gunther_forcing}, 
we were able to define this relation and prove its properties relatively easily.

@@

Where \texttt{"wellordered(C, A, r)"} denotes that 
the relation \texttt{"r"} is a well-ordering of the set \texttt{"A"} in the class \texttt{"C"}, 
and \texttt{"\#\#"} is an operator that converts a set to a class.
These are defined by \cite{paulson_AC_consistency}.
Specifically, the definition of \texttt{"wellordered"} is as follows
\footnote{
Note that we can show that 
if \texttt{"wellordered(N, A, r)"} holds for an element \texttt{"A"} in a class \texttt{"N"}, which is a model of ZF,
then \texttt{"r"} is irreflexive and assymetric, i.e., a strict total order.
}:
\begin{isaframe}
\begin{minted}[breaklines]{isabelle}
definition transitive_rel :: "[i⇒o,i,i]⇒o" where
    "transitive_rel(M,A,r) ≡ 
        ∀x[M]. x∈A ⟶ (∀y[M]. y∈A ⟶ (∀z[M]. z∈A ⟶ 
                          ⟨x,y⟩∈r ⟶ ⟨y,z⟩∈r ⟶ ⟨x,z⟩∈r))"
definition linear_rel :: "[i⇒o,i,i]⇒o" where
    "linear_rel(M,A,r) ≡ 
        ∀x[M]. x∈A ⟶ (∀y[M]. y∈A ⟶ ⟨x,y⟩∈r | x=y | ⟨y,x⟩∈r)"

@@

\chapter{Conclusion} \label{chap:conclusion}

We formalized the relative consistency proof of $\neg$AC with ZF
on the basis of the work by \cite{paulson_AC_consistency} and \cite{gunther_forcing}.
Specifically, assuming the existence of a c.t.m.\ of ZF, 
we constructed a model of ZF+$\neg$AC known as the Basic Cohen model by forcing. 
To achieve this, we formalized the definition of symmetric extensions and the related lemmas.

@@

Through the formalization, we felt particularly deficient of 
functionality to efficiently formalize “object-level” arguments in axiomatic set theory.

In our proof, we needed to construct internalized formulas 
representing most of the concepts that belong to the ground model. 
This was to verify that the defined sets truly belong to the ground model 
and to enable application of the separation and replacement axioms of the ground model for formulas that include these concepts.

For constructing these formulas we utilized the sythesis method by \cite{gunther_forcing};
however, this method could not be applied to complex formulas.
When applied, we also needed to reprove properties of derived formulas in a more usable form.

@@

@@

Additionally, we used the c.t.m.\ approach, and the formalized argument only extends to the construction of a model of ZF+$\neg$AC,
assuming the existence of a c.t.m.\ of ZF.
We did not formalize the justification of the c.t.m.\ approach itself.
However, it is also desirable to formalize this part.

The main obstacle to formalizing this part lies in the reflection principle. 
The reflection principle for propositions of Isabelle 
was proved by \cite{paulson_reflection}; 
however, since we are dealing with propositions formalized in the form of internalized formulas, 
we need a reflection principle for such formulas.

There are several approaches to address this. 
One approach is to use metaprogramming, 
which involves writing a program to automatically generate proofs 
for specific instances of the reflection principle. 
In this case, only a finite number of instances of the reflection principle can be utilized, 
and the finite set of logical formulas as parameters for each instance must be explicitly constructed. 
This means that the reflection principle cannot be applied to finite sets of logical formulas obtained non-constructively.

Another approach is to prove a general reflection principle for internalized formulas. 
This would require discussing "object-level within object-level", that is, 
a further layer of encoded set theory within models of ZF or ZFC. Since even "object-level" arguments, 
such as those handled in this formalization, are already sufficiently complex, 
this approach would be difficult unless the aforementioned automation is achieved.

The final approach is to use higher-order logic. 
In higher-order logic, it would be possible to prove the reflection principle for internalized formulas 
without engaging in the complex discussions of "object-level within object-level". 
However, this would require using a proof assistant other than Isabelle/ZF. 
Since the accumulated definitions and lemmas in Isabelle/ZF would not be available, 
it would be necessary to re-formalize set-theoretic concepts such as forcing.

\subsubsection{Future Work}
The automation of "object-level" arguments and the formalization of the validity of the c.t.m.\ approach mentioned above are interesting future work. 
Additionally, it would be interesting to formalize more advanced topics on symmetric extensions, 
as seen in the latter parts of \cite{karagila}. 
Furthermore, formalizing the fact that the consistency of an axiomatic system $S$ does not imply the existence of a c.t.m.\ of $S$, 
as well as alternative approaches to forcing, 
such as those using modal logic, 
and formalizing the connections between different approaches to forcing, 
are also intriguing topics for future exploration.
\printbibliography
\end{document}
